---
title: "Which Weather Events Affect Public Health And Economy The Most According To NOAA Storm Database (1950 - 2011)"
author: "Daren Thomas"
date: "5 May 2016"
output: html_document
---

## Synopsis

This Report uses the Storm Database of the U.S. National Oceanic and Atmospheric Administration (NOAA) to identify those weather events that create the most harm to public health and the economy. Harm to public health is shown as total fatalities by event type and total injuries by event type. Harm to the economy looks at property damage by event type as well as damage to crops by event type. Data from 1950 to 2011 are considered. An approach to automatically cleaning event type labels in the data is attempted with limited success.

The most harmful event type for public health is `TORNADO`. `FLOOD` creates the most property damage, followed by `HURRICANETYPHOON` which also happens to be the most harmful event regarding crop damage.

## Data Processing

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can [download the file from the course web site](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2): 

```{r cache=TRUE} 
noaa.url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
noaa.bz2 = "storm-data.csv.bz2"
if (! file.exists(noaa.bz2)) {
    download.file(url=noaa.url, destfile=noaa.bz2)
}
noaa <- read.csv(noaa.bz2)
```

After reading, we check the first few rows (there are `r nrow(noaa)` ) in the dataset:

```{r}
dim(noaa)
```

```{r}
head(noaa)
```

Also, I am interested in the types of the fields:

```{r}
str(noaa)
```

It looks like the variables I'm most interested in are:

- `EVTYPE`
- `FATALITIES`
- `INJURIES`
- `PROPDMG`
- `CROPDMB`

### Processing of Field `EVTYPE`
```{r}
require(dplyr)
```

```{r}
nrow(distinct(select(noaa, EVTYPE)))
```

The field `EVTYPE` has `r nrow(distinct(select(noaa, EVTYPE)))` distinct values. Going through them, it seems some are summaries. Others have labels that start with spaces. There are some inconsistencies with uppercase/lowercase - let's just move everything to uppercase. I am also going to check for typos.

First, remove surrounding whitespace on event types and uppercase everything:

```{r}
noaa$EVTYPE2 <- trimws(noaa$EVTYPE)
noaa$EVTYPE2 <- toupper(noaa$EVTYPE2)
nrow(distinct(select(noaa, EVTYPE2)))
```

The new field `EVTYPE2` has `r nrow(distinct(select(noaa, EVTYPE2)))`, less than `EVTYPE`, so we are making progress.

Next, remove any summary rows, and while we are at it, any NA rows (those with value `?`):

```{r}
noaa2 <- filter(noaa, !grepl('summary', noaa$EVTYPE2, ignore.case=TRUE))
noaa2 <- filter(noaa2, EVTYPE2 != '?')
nrow(distinct(select(noaa2, EVTYPE2)))
```

The new dataset `noaa2` has only `r nrow(distinct(select(noaa2, EVTYPE2)))` distinct event types.

Sorting the field has the advantage that it makes it easy to spot typos in the data (such as `AVALANCE` and `AVALANCHE`)

```{r}

# NOTE: the code below will generate a long list of event types
evtypes <- distinct(select(noaa2, EVTYPE2)) %>% arrange(EVTYPE2)
head(evtypes)
```

Next, we slog through this list, but we're lazy, so we are going to compare using `adist` and assume that a distance < 2 is a typo. We will update the original event type with the match that has the most entries, assuming that the typo is less common than the original value. This does create a problem of `THUNDERSTORM WIND 50` being lumped togeter with `THUNDERSTORM WIND 60`. So, for this analysis, I am going to remove all numbers from event types as well as numbers that start with a letter (`F0`, `G50` etc.) and parenthesis too.

```{r}
# remove numbers with potential leading letter from event types
noaa2$EVTYPE3 <- gsub('[A-Z]?[0-9]+', '', noaa2$EVTYPE2)

# remove ' MPH'
noaa2$EVTYPE3 <- sub(' MPH', '', noaa2$EVTYPE3)

# remove any non-letter character from event types
noaa2$EVTYPE3 <- gsub('[^A-Z]', '', noaa2$EVTYPE3)
```

Next we update similar event types, using the most common one as the probably correct one:
(NOTE: I didn't just create the whole matrix/vectorize, because R Studio kept crashing... therefore we loop!)

```{r}
evtypes <- distinct(select(noaa2, EVTYPE3)) %>% arrange(EVTYPE3)
for (row in 1:(nrow(evtypes)-1)) {
    left <- evtypes[row, 1]
    right <- evtypes[row+1, 1]
    distance <- adist(left, right)[1, 1]
    if (distance < 2) {
        nleft <- length(noaa2[noaa2$EVTYPE3 == left, 'EVTYPE3'])
        nright <- length(noaa2[noaa2$EVTYPE3 == right, 'EVTYPE3'])
        if (nleft > nright) {
            noaa2$EVTYPE3[noaa2$EVTYPE3 == right] <- left
            #print(c(right, length(noaa2[noaa2$EVTYPE3 == right, 'EVTYPE3'])))
        }
        else {
            noaa2$EVTYPE3[noaa2$EVTYPE3 == left] <- right
            #print(c(left, length(noaa2[noaa2$EVTYPE3 == left, 'EVTYPE3'])))
        }
    }
}
evtypes <- distinct(select(noaa2, EVTYPE3)) %>% arrange(EVTYPE3)
```

It turns out that this approach is only a first stab - we'd have to actually define a smaller set of event types and a mapping to those types. For the purposes of this analysis, I think we have done enough.

Let us look at the other fields we'd like to analyse:

```{r}
noaa2$FATALITIES <- as.numeric(noaa2$FATALITIES)
noaa2$INJURIES <- as.numeric(noaa2$INJURIES)
noaa2$PROPDMG <- as.numeric(noaa2$PROPDMG)
noaa2$CROPDMG <- as.numeric(noaa2$CROPDMG)

noaa2$FATALITIES <- as.numeric(noaa2$FATALITIES)
noaa2$INJURIES <- as.numeric(noaa2$INJURIES)
```

Going through the documentation, it seems as if the `PROPDMGEXP` and `CROPDMGEXP` fields contain an exponend to multiply the `PROPDMG` and `CROPDMG` values with. I need to clean this up since a couple of conventions are used:

```{r}
# first all uppercase
noaa2$PROPDMGEXP <- as.character(noaa2$PROPDMGEXP)
noaa2$PROPDMGEXP <- toupper(noaa2$PROPDMGEXP)
noaa2$PROPDMGEXP <- trimws(noaa2$PROPDMGEXP)

# replace `?` and `-` and `+` and ` ` with `0`
noaa2$PROPDMGEXP[noaa2$PROPDMGEXP %in% c('?', '-', '', '+')] <- '0'

# replace K, M, B, H with 3, 6, 9, 2 respectively
noaa2$PROPDMGEXP[noaa2$PROPDMGEXP == 'K'] <- '3'
noaa2$PROPDMGEXP[noaa2$PROPDMGEXP == 'M'] <- '6'
noaa2$PROPDMGEXP[noaa2$PROPDMGEXP == 'B'] <- '9'
noaa2$PROPDMGEXP[noaa2$PROPDMGEXP == 'H'] <- '2'
noaa2$PROPDMGEXP <- as.numeric(noaa2$PROPDMGEXP)


# create new field with multiplied values
noaa2 <- mutate(noaa2, property.damage = PROPDMG * (10 ^ PROPDMGEXP))

# repeat for crop damage
# first all uppercase
noaa2$CROPDMGEXP <- as.character(noaa2$PROPDMGEXP)
noaa2$CROPDMGEXP <- toupper(noaa2$PROPDMGEXP)
noaa2$CROPDMGEXP <- trimws(noaa2$PROPDMGEXP)

# replace `?` and `-` and `+` and ` ` with `0`
noaa2$CROPDMGEXP[noaa2$CROPDMGEXP %in% c('?', '-', '', '+')] <- '0'

# replace K, M, B, H with 3, 6, 9, 2 respectively
noaa2$CROPDMGEXP[noaa2$CROPDMGEXP == 'K'] <- '3'
noaa2$CROPDMGEXP[noaa2$CROPDMGEXP == 'M'] <- '6'
noaa2$CROPDMGEXP[noaa2$CROPDMGEXP == 'B'] <- '9'
noaa2$CROPDMGEXP[noaa2$CROPDMGEXP == 'H'] <- '2'
noaa2$CROPDMGEXP <- as.numeric(noaa2$CROPDMGEXP)


# create new field with multiplied values
noaa2 <- mutate(noaa2, crop.damage = CROPDMG * (10 ^ CROPDMGEXP))

# next, let's just drop the fields we're not interested in:
noaa3 <- select(noaa2, EVTYPE3, INJURIES, FATALITIES, property.damage, crop.damage)
```

## Results

I am trying to answer these two questions:

1. Across the United States, which types of events (as indicated in the ð™´ðš…ðšƒðšˆð™¿ð™´ variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?


### Most harmful with respect to population health

Let us say, population health is impaired by fatalities and injuries. So, which events have the highest count of fatalities?

```{r}
fatal <- noaa3 %>% group_by(EVTYPE3) %>% summarize(total.fatalities=sum(FATALITIES)) %>%  
    arrange(desc(total.fatalities))
top.fatal <- head(fatal, n=10)
top.fatal
```

Viewing the `TSTMWIND` event type, we can see shortcomings of the data cleaning step - future research could try to do a better job here, but a quick glance at the numbers suggests that the top three (TORNADO, EXCESSIVEHEAT and FLASHFLOOD) will not be altered.

What events cause the most injuries? We can use the same technique here:

```{r}
injuries <- noaa3 %>% group_by(EVTYPE3) %>% summarize(total.injuries=sum(INJURIES)) %>%  
    arrange(desc(total.injuries))
top.injuries <- head(injuries, n=10)
top.injuries
```

Here, again, shortcommings of my data cleaning step show up. It is still clear, though, that `TORNADO` is the big bad event!

Let's show this graphically:

```{r}
require(ggplot2)

par(mfrow=c(1, 2), mar = c(12, 4, 2, 2.5), cex=0.7)
barplot(top.fatal$total.fatalities, names.arg=top.fatal$EVTYPE3, ylab='Fatalities', las=3, 
        main='Fatalities by Event type')
barplot(top.injuries$total.injuries, names.arg=top.injuries$EVTYPE3, ylab='Injuries', las=3, 
        main='Injuries by Event type')
```

### Most harmful with respect to the economy

Next, we look at the cost of these events to the economy. We look at both property damage and damage to crops:

```{r}
property <- noaa3 %>% group_by(EVTYPE3) %>% summarize(total.property.damage=sum(property.damage)) %>%  
    arrange(desc(total.property.damage))
top.property <- head(property, n=10)
top.property
```

```{r}
crop <- noaa3 %>% group_by(EVTYPE3) %>% summarize(total.crop.damage=sum(crop.damage)) %>%  
    arrange(desc(total.crop.damage))
top.crop <- head(crop, n=10)
top.crop
```

```{r}
require(ggplot2)

par(mfrow=c(1, 2), mar = c(12, 4, 2, 2.5), cex=0.7)
barplot(top.property$total.property.damage / 1000, names.arg=top.property$EVTYPE3, ylab='Property Damage (1000 $)', las=3, 
        main='Property Damage by Event type')
barplot(top.crop$total.crop.damage / 1000, names.arg=top.crop$EVTYPE3, ylab='Crop Damage (1000 $)', las=3, 
        main='Crop Damage by Event type')
```

We can see that `HURRICANE` and `HURRICANETYPHOON` are both the most harmful event types for crops. These probably could  even be lumped into a single event type `HURRICANE` (see data cleaning section).

We can also see that `FLOOD` causes the most property damage, with `HURRICANETYPHOON` and `TORNADO` among the top 3 too!

I would be interested in seeing how much these numbers change if I had time to reassign each event type to a smaller list (say, 20-30 types) by doing a more rigorous data cleaning. I do not really think the #1's will change, though.